``Interestingness'' in Mathematical Theory Exploration

### 2018-08-20 PhDThesis/background_haskell.tex
Lit review and background

Whilst \S \ref{sec:haskell} gave some brief background on Haskell, little
explanation was given for why we chose this language rather than, for example
Coq or ACL2 (for which recurrent clustering algorithms already exist), or a more
widely used language like Java. Here we discuss the relevant language features
from a high-level, which motivated our choice:

@@

\item{Statically Typed}: Expression are constrained by \emph{types}, which can
  be used to eliminate unwanted combinations of values, and hence reduce search
  spaces; \emph{static} types can be deduced syntactically, without having to
  execute the code.

@@

Together, these features make Haskell code highly structured, amenable to
logical analysis and subject to many algebraic laws. However, as mentioned with
regards to type classes, Haskell itself is incapable of expressing or enforcing
these laws (at least, without difficulty \cite{lindley2014hasochism}). This
reduces the incentive to manually discover, state and prove theorems about
Haskell code, e.g. in the style of interactive theorem proving, as these results
may be invalidated by seemingly innocuous code changes. This puts Haskell in a
rather special position with regards to the discovery of interesting theorems;
namely that many discoveries may be available with very little work, simply
because the code's authors are focused on \emph{software} development rather
than \emph{proof} development. The same cannot be said, for example, of ITP
systems; although our reasoning capabilities may be stronger in an ITP setting,
much of the ``low hanging fruit'' will have already been found through the
user's dedicated efforts, and hence theory exploration would be unlikely to
discover unexpected properties.

@@

\item The existence of theory exploration systems such as \hspec{}, and related
  tools which we may be able to re-use, including conjecture generators like
  \qspec{}; counterexample finders like \qcheck{}, \textsc{SmallCheck} and
  \textsc{SmartCheck}; theorem provers like \textsc{Hip}
  \cite{rosen2012proving}; and other related testing and term-generating
  systems like \textsc{MuCheck} \cite{le2014mucheck}, \textsc{MagicHaskeller}
  \cite{katayama2011magichaskeller} and \textsc{Djinn}
  \cite{augustsson2005djinn}.

>>>>

We mostly focus our attention on the Haskell programming language, both as an
implementation vehicle and as our representation of functions, properties, etc.
to explore. We choose Haskell since it combines formal, logical underpinnings
which aid reasoning (compared to more popular languages like Java or C), yet it
is still popular enough to sustain a rich ecosystem of tooling and a large
corpus of existing code (compared to more formal languages like Coq or
Isabelle).

The following features make Haskell especially useful for our purposes, although
many are also present in other languages such as StandardML and Coq (which we
also use, but only when needed for compatibility with prior work):

@@

\item{Statically Typed}: Expressions are constrained by \emph{types}, which can
  be used to eliminate unwanted combinations of values, and hence reduce search
  spaces; \emph{static} types can be deduced syntactically, without having to
  execute the code.

@@

Together, these features make Haskell code highly structured, amenable to
logical analysis and subject to many algebraic laws. However, as mentioned with
regards to type classes, Haskell itself is incapable of expressing or enforcing
these laws (at least, without difficulty \cite{lindley2014hasochism}). This
reduces the incentive to manually discover, state and prove theorems about
Haskell code, e.g. in the style of interactive theorem proving, as these results
may be invalidated by seemingly innocuous code changes. This puts Haskell in a
rather special position with regards to the discovery of interesting theorems;
namely that many discoveries may be available with very little work, simply
because the code's authors are focused on \emph{software} development rather
than \emph{proof} development. The same cannot be said, for example, of ITP
systems; although our reasoning capabilities may be stronger in an ITP setting,
much of the ``low hanging fruit'' will have already been found through the
user's dedicated efforts, and hence theory exploration would be less likely to
discover unexpected properties.

@@

\item The existence of theory exploration systems such as \hspec{}, and related
  tools which we can to re-use, including conjecture generators like \qspec{}
  and \speculate{}; counterexample finders like \qcheck{}, \textsc{SmallCheck},
  \textsc{SmartCheck}, \textsc{LeanCheck}, \textsc{Hedgehog}, etc.; theorem
  provers like \textsc{Hip} \cite{rosen2012proving}; and other related testing
  and term-generating systems like \textsc{MuCheck} \cite{le2014mucheck},
  \textsc{MagicHaskeller}~\cite{katayama2011magichaskeller} and
  \textsc{Djinn}~\cite{augustsson2005djinn}.

### 2018-08-20 PhDThesis/litreview.tex
Lit review and background

\input{litreview_mathhistory.tex}
\input{litreview_interestingness.tex}
\input{litreview_theoryformation.tex}
\input{litreview_theoryexploration.tex}
\input{litreview_premiseselection.tex}

>>>>

The automation of mathematical tasks and reasoning has been pursued since at
least the time of mechanical calculators like the Pascaline~\cite{d'ocagne}. A
recurring theme in these efforts is the separation between those undertaken by
mathematicians like Pascal and Babbage~\cite{bowden}, and those of engineers
such as M\"uller~\cite[p. 65]{lindgren}. This pattern continues today, with the
tasks we are concerned with (automatically constructing and evaluating concepts,
conjectures, theorems, axioms, examples, etc.) being divided into two main
fields: Mathematical Theory Exploration (MTE)~\cite{buchberger:06} (also
sometimes prefaced with ``Computer-Aided'', ``Automated'' or
``Algorithm-Supported''), which is championed by mathematicians such as
Buchberger~\cite{buchberger}; and Automated Theory Formation
(ATF)~\cite{lenat:77,colton:book}, pursued by AI researchers including Lenat.
Other related terms include ``Automated Mathematical
Discovery''~\cite{epstein:91,colton2000notion,esarm2008},
``Concept Formation in Discovery Systems''~\cite{haase}, and
``Automated Theorem Discovery''~\cite{roy}.

Such a plethora of terminology can mask similarities and shared goals between
these fields. Even notable historical differences, such as the emphasis of MTE
on user-interaction and mathematical domains, in contrast to the full automation
and more general applications targeted by ATF, are disappearing in recent
implementations.

\subsection{Theory Formation}

An important historical implementation of ATF is Lenat's AM (Automated
Mathematician) system. Unlike prior work, such as
Meta-Dendral~\cite{buchanan:75} and those described in~\cite{winston}, AM aims
to be a general purpose mathematical discovery system, designed to both
construct new concepts and conjecture relationships between them. AM is a
rule-based system which represents knowledge using a frame-like scheme, enlarges
its knowledge base via a collection of heuristic rules, and controls the firing
of these rules via an agenda mechanism.

The performance of AM was evaluated based on its generality (performance in new
domains) and how finely-tuned various aspects of the program are (the agenda,
the interaction of the heuristics, etc). Most of this evaluation was
qualitative, and has subsequently been criticised~\cite[chap.~13]{colton:book}.
In their case study in methodology, Ritchie and Hanna found a large discrepancy
between the theoretical claims made of AM and the implemented
program~\cite{ritchie1984case}; for example, AM ``invented'' natural numbers
from sets, but did so using a heuristic specifically designed to make this
connection.

The successor of AM is Eurisko, a discovery system intended to be useful in more
general domains than just mathematical theory formation. Despite claiming some
early successes % FIXME Traveller TCS
work on Eurisko was put on hold due to a lack of machine-readable data about
real world systems for it to work with % FIXME Find something quotable
. This lead to the Cyc project, in an attempt to encode such knowledge in a
semantically rich and consistent form. Cyc is an ongoing project, with parts of
its database made available via OpenCyc.

\subsection{Theory Exploration}

The prototypical implementation of MTE is the Theorema system of Buchberger and
colleagues~\cite{buchberger,buchberger2016theorema}, which also places a strong
emphasis on user interface and output presentation. Theory exploration in the
Theorema system involves the user formalising their definitions in a consistent,
layered approach; such that reasoning algorithms can exploit this structure in
subsequent proofs, calculations, etc. The potential of this strategy was
evaluated by illustrating the automated synthesis of Buchberger's own Gr\"obner
bases algorithm~\cite{buchberger:04}.

A similar ``layering'' approach is found in the IsaScheme system of
Monta{\~n}o-Rivas \etal{}~\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, which
has also been quantitatively compared against IsaCoSy and HipSpec using
precision/recall analysis~\cite{claessen2013automating}. The name comes from its
embedding in the Isabelle proof assistant and its use of ``schemes'':
higher-order formulae which can be used to generate new concepts and
conjectures. Variables within a scheme are instantiated automatically and this
drives the invention process. For example, the concept of ``repetition'' can be
encoded as a scheme, and instantiated with existing encodings of zero, successor
and addition to produce a definition of multiplication. The same scheme can be
instantiated with this new multiplication function to produce exponentiation.

IsaCoSy and QuickSpec (the conjecture generation component of HipSpec) are
described in more detail in $\S$\ref{sec:existing-tools}, since these are the
tools we chose to evaluate and compare for $\S$\ref{sec:application}. QuickSpec
has since evolved to version 2~\cite{smallbone2017quick}, which replaces the
distinct enumeration and testing steps with a single, iterative algorithm
similar to that of IsaCoSy. Generated conjectures are fed into a Knuth-Bendix
completion algorithm to form a corresponding set of rewrite rules. As
expressions are enumerated, they are simplified using these rules and discarded
if equal to a known expression. If not, QuickCheck tests whether the new
expression can be distinguished from the known expressions through random
testing: those which can are added to the set of known expressions. Those which
cannot be distinguished are conjectured to be equal, and the rewrite rules are
updated.

QuickSpec has also inspired another MTE tool for Haskell called
Speculate~\cite{braquehais2017speculate}, which operates in a similar way but
also makes use of the laws of total orders and Boolean algebra to conjecture
\emph{in}equalities and conditional relations between expressions.

Another notable MTE implementation, distinct from those based in Isabelle and
Haskell, is the MATHsAiD project (Mechanically Ascertaining Theorems from
Hypotheses, Axioms and Definitions)~\cite{roy}. Unlike the tools above, which
generate \emph{conjectures} that may later be sent to automated provers,
MATHsAiD directly generates \emph{theorems}, by making logically valid
inferences from a given set of axioms and definitions. Evaluation of the
interestingness of these theorems was performed qualitatively by the system's
developer, which highlights how these tools could benefit from the availability
of an objective, repeatable, quantitative method of evaluation and comparison
such as ours.

\input{litreview_staticanalysis.tex}
\input{litreview_premiseselection.tex}
\input{litreview_interestingness.tex}

### 2018-08-20 PhDThesis/litreview_premiseselection.tex
Lit review and background

\subsubsection{Relevance Filtering}
\label{sec:relevance}

@@

The combinatorial nature of formal systems causes many proof search methods,
such as resolution, to have exponential complexity
\cite{haken1985intractability}; hence even a modest size increase can turn a
trivial problem into an intractable one. Finding efficient alternatives for such
algorithms, especially those which are NP-complete (e.g. determining
satisfiability) or co-NP-complete (e.g. determining tautologies), seems
unlikely, as it would imply progress on the famously intractable open problems
of $\text{P} = \text{NP}$ and $\text{NP} = \text{co-NP}$. On the other hand, we
can turn this difficulty around: a modest \emph{decrease} in size may turn an
intractable problem into a solvable one. We can ensure that the solutions to
these reduced problems coincide with the original if we only remove
\emph{redundant} information. This leads to the idea of \emph{relevance
  filtering}.

@@

The initial approach, known as \textsc{MePO} (from \emph{Meng-Paulson}
\cite{meng2009lightweight}), gives each clause a score based on the proportion
$m / n$ of its symbols which are ``relevant'' (where $n$ is the number of
symbols in the clause and $m$ is the number which are relevant). Initially, the
relevant symbols are those which occur in the goal, but whenever a clause is
found which scores more than a particular threshold, all of its symbols are then
also considered relevant. There are other heuristics applied too, such as
increasing the score of user-provided facts (e.g. given by keywords like
\texttt{using}), locally-scoped facts, first-order facts and rarely-occuring
facts. To choose $r$ relevant clauses for an ATP invocation, we simply order the
clauses by decreasing score and take the first $r$ of them.

Recently, a variety of alternative algorithms have also been investigated,
including:

\begin{description}
\item{\textsc{MaSH}}: Machine Learning for SledgeHammer
  \cite{kuhlwein2013mash}. The distinguishing feature of \textsc{MaSH} is its
  use of ``visibility'', which is essentially a dependency graph of which
  theorems were used in the proofs of which other theorems; although theorems
  are represented as abstract sets of features. To select relevant clauses for a
  goal, the set of clauses which are visible from the goal's components is
  generated; this is further reduced by (an efficient approximation of) a naive
  Bayes algorithm.

\item{\textsc{MOR}}: \emph{Multi-output ranking} uses a support vector machine
  (SVM) approach for selecting relevant axioms from the Mizar Mathematical
  Library for use by the Vampire ATP system \cite{alama2014premise}. \iffalse
  TODO: describe the kernel, as that's the interesting bit \fi It compares
  favourably to \textsc{SNoW} and \textsc{SInE}.

  % TODO:
\item{\textsc{SInE}}
\item{\textsc{BliStr}}
\item{\textsc{HOLyHammer}}
\item{\textsc{MoMM}}
\item{\textsc{SNoW}}
\item{\textsc{MPTP 0.2}}
\item{\textsc{MaLARea}}
\item{\textsc{MaLARea SG1}}

\end{description}

>>>>

\subsection{Relevance Filtering}
\label{sec:relevance}

@@

One fruitful application of machine learning methods to formal mathematics is
\emph{relevance filtering}, which is used for \emph{premise selection} in
automated theorem provers.

@@

Sledgehammer's initial approach, known as \textsc{MePO} (from
\emph{Meng-Paulson} \cite{meng2009lightweight}), gives each clause a score based
on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is
the number of symbols in the clause and $m$ is the number which are relevant).
Initially, the relevant symbols are those which occur in the goal, but whenever
a clause is found which scores more than a particular threshold, all of its
symbols are then also considered relevant. There are other heuristics applied
too, such as increasing the score of user-provided facts (e.g. given by keywords
like \texttt{using}), locally-scoped facts, first-order facts and
rarely-occuring facts. To choose $r$ relevant clauses for an ATP invocation, we
simply order the clauses by decreasing score and take the first $r$ of them.

Recently, a variety of alternative algorithms have also been investigated,
dubbed \textsc{MaSH}, \textsc{SInE}, \textsc{BliStr}, \textsc{HOLyHammer},
\textsc{MoMM}, \textsc{SNoW}, \textsc{MPTP 0.2}, \textsc{MaLARea} and
\textsc{MaLARea SG1}. Each one combines an ``off the shelf'' machine learning
algorithm with a custom feature extraction approach, is trained using a corpus
of existing theorems and their proofs, and evaluated by trying to re-prove a
selection of unseen theorems from the corpus (i.e. given only the theorem
statement and possible lemmas, not the existing proof).

For example the Machine Learning for SledgeHammer (\textsc{MaSH}) system
defines a notion of ``visibility''~\cite{kuhlwein2013mash}. This is essentially
a dependency graph of which theorems were used in the proofs of which other
theorems (theorems are represented as abstract sets of features, rather than
e.g. sentences or syntax trees). To select relevant clauses for a goal, the
set of clauses which are ``visible'' from the goal's components is generated;
this is further reduced by (an efficient approximation of) a naive Bayes
algorithm.

The Multi-output ranking (\textsc{MOR}) implementation uses a support vector
machine (SVM) to select relevant axioms from the Mizar Mathematical Library for
use by the Vampire ATP system~\cite{alama2014premise}.
%TODO: describe the kernel, as that's the interesting bit.

The premise selection problem for theorem proving is related to the signature
selection problem for theory exploration, in that both are trying to pick a
small number of terms (depending on the speed of the subsequent algorithm and
how long the user is willing to wait) from a large set, to optimise the
outcome of that algorithm (e.g. finding a proof quickly; or finding many
interesting relationships quickly). The major difference is that premise
selection has a particular ``target'' theorem statement, against which to
compare each term (e.g. for the ``relevance'' of their sub-terms). This is not
the case for theory exploration, so we must consider the interactions of each
term with all of the others when deciding what to select.
